Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-03 mr-review.Rmd
added to 1-instrument, 1-outcome methods

Great interest has developed in inferring causal relationships between complex 
traits in observational human genetics studies. Throughout the manuscript, we 
use the term "complex trait" to refer to a trait that doesn't appear to be 
inherited in a Mendelian fashion. Typically, such complex traits such as gene expression levels
are inherited in a polygenic manner, with many genes influencing the trait.

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via mendelian randomization. A genome-wide association study 
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
SNP is a genetic marker defined by nucleotide base identity at a single genomic position.

Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
appearance of the first GWAS.

@@

We begin our discussion of Mendelian randomization by considering the simplest case, where there is a single genetic marker and a single outcome variable. Mendelian randomization with one instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
whether variation in the genotypes at the genetic marker causes variation in the 
"outcome" variable, which is typically a second complex trait. When the genetic marker 
is a single nucleotide polymorphism on an autosome, the cohort is partitioned into three 
classes, two homozygote classes and a single heterozygote class, based on genotypes 
for that marker. When the three assumptions of mendelian randomization are satisfied, 
estimation of the causal effect of the exposure on the outcome is possible.

@@

Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer 
causal effects in a MR study. Careful evaluation of the assumptions is needed in MR studies. Briefly, 
the first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies.

@@

The second assumption, often termed the exclusion restriction, can also be stated as 
the need for the genetic marker to affect the outcome only through the exposure. In a 
causal diagram, this means that the only path from the marker to the outcome is that 
containing the exposure.

The third assumption, sometimes called the ____, ensures that ___.

@@

>>>>

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Two-sample 
Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects. Importantly, the SNP instruments are measured in both sets of subjects.

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
SNP is a genetic marker defined by nucleotide base identity at a single genomic position.

```{r, article-counts, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
knitr::include_graphics("figures/mr-gwas-counts-scatter.png")
```

Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
appearance of the first GWAS (Figure \@ref(fig:article-counts)).

Recent developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is an abundance of freely available GWAS summary data [@buniello2019nhgri]. GWAS summary statistics typically consist of an identifier and a p-value for every SNP in the data set [@buniello2019nhgri].

@@

We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single genetic marker and a single outcome variable. Mendelian randomization with one instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
whether variation in the genotypes at the genetic marker causes variation in the 
"outcome" variable, which is typically a second complex trait. When the genetic marker 
is a SNP on an autosome, the cohort is partitioned into three 
classes, two homozygote classes and a single heterozygote class, based on genotypes 
for that marker. When the three assumptions of mendelian randomization are satisfied, 
estimation of the causal effect of the exposure on the outcome is possible.

@@

Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer 
causal effects in an observational MR study. Careful evaluation of the assumptions is needed. Briefly, 
the first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies.

@@

The second assumption, often termed the exclusion restriction assumption, can also be stated as 
the need for the genetic marker to affect the outcome only through the exposure. In a 
causal diagram, this means that the only path from the marker to the outcome is that 
containing the exposure.

The third assumption, sometimes called the , ensures that .

@@

A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes). The resulting 
fitted values for the exposure variables are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient obtained from the second regression analysis.
